Quotients of the magmatic operad:\\ lattice structures and convergent rewrite systems

### 2018-09-09 CAs_operads.tex
Dessins partie 3 (2/2)
Relecture partie 3 (avant prop. 3.2.7)

\begin{Lemma} \label{lem:surjective_morphisms_CAs}
    Let $\gamma$ and $\gamma'$ be two positive integers. If there exists
    an operad morphism $\varphi:\CAs{\gamma'} \to \CAs{\gamma}$, then it
    is surjective and unique.
\end{Lemma}
\begin{proof}
    The operad $\CAs{\gamma'}$ is generated by one binary generator
    $[\Product]_{\CongrCAs{\gamma'}}$, which is the image of the binary
    generator $\Product$ of $\Mag$ in $\CAs{\gamma'}$. Hence, $\varphi$
    is entirely determined by the image
    $\varphi\left([\Product]_{\CongrCAs{\gamma'}}\right)$. Moreover,
    $\varphi([\Product]_{\CongrCAs{\gamma'}})$ has to be of arity $2$ in
    $\CAs{\gamma}$, so that we necessarily have
    \begin{math}
        \varphi\left([\Product]_{\CongrCAs{\gamma'}}\right)
        =
        [\Product]_{\CongrCAs{\gamma}}.
    \end{math}
    Hence, if $\varphi$ exists, it is the unique operad morphism from
    $\CAs{\gamma'}$ to $\CAs{\gamma}$ determined by the image of
    $[\Product]_{\CongrCAs{\gamma'}}$. In this case,
    $[\Product]_{\CongrCAs{\gamma}}$ being in the image of $\varphi$,
    the latter is surjective.
\end{proof}
\medbreak

@@

\begin{Lemma} \label{lem:morphism_CAs}
    Let $\gamma$ and $\gamma'$ be two positive integers. There exists an
    operad morphism $\varphi:\CAs{\gamma'} \to \CAs{\gamma}$ if and only
    if
    \begin{math}
      \LComb{\gamma'} \CongrCAs{\gamma} \RComb{\gamma'}.
    \end{math}
\end{Lemma}
\begin{proof}
    Assume that $\varphi:\CAs{\gamma'} \to \CAs{\gamma}$ is an operad
    morphism. By Lemma~\ref{lem:surjective_morphisms_CAs}, $\varphi$
    satisfies
    \begin{math}
        \varphi\left(\left[\Tfr\right]_{\CongrCAs{\gamma'}}\right)
        = \left[\Tfr\right]_{\CongrCAs{\gamma}}
    \end{math}
    for any binary tree $\Tfr$. Since
    \begin{math}
        \varphi\left(
        \left[\LComb{\gamma'}\right]_{\CongrCAs{\gamma'}}
        \right)
        =
        \varphi\left(
        \left[\RComb{\gamma'}\right]_{\CongrCAs{\gamma'}}
        \right),
    \end{math}
    we have
    \begin{math}
        \left[\LComb{\gamma'}\right]_{\CongrCAs{\gamma}}
        =
        \left[\RComb{\gamma'}\right]_{\CongrCAs{\gamma}},
    \end{math}
    which is equivalent to the fact that
    \begin{math}
        \LComb{\gamma'} \CongrCAs{\gamma} \RComb{\gamma'}.
    \end{math}
    \smallbreak

@@

\begin{Proposition}\label{prop:poset_CAs}
    The binary relation $\OrdCAs$ is a partial order relation on
    $\CAsAll$.
\end{Proposition}
\begin{proof}
    The binary relation $\OrdCAs$ is reflexive since there exists the
    identity morphism of $\CAs{\gamma}$ for every positive integer
    $\gamma$. Let us now assume that
    $\varphi:\CAs{\gamma''} \to \CAs{\gamma'}$ and
    $\psi:\CAs{\gamma'} \to \CAs{\gamma}$ are two operad morphisms. By
    Lemma~\ref{lem:morphism_CAs}, one has
    $\LComb{\gamma''} \CongrCAs{\gamma'} \RComb{\gamma''}$ and
    $\LComb{\gamma'} \CongrCAs{\gamma} \RComb{\gamma'}$. This implies
    that $\CongrCAs{\gamma}$ is coarser than $\CongrCAs{\gamma'}$ and
    that $\CongrCAs{\gamma'}$ is coarser that $\CongrCAs{\gamma''}$. Now,
    again by Lemma~\ref{lem:morphism_CAs}, $\psi \circ \varphi$ is an
    operad morphism. Hence, $\OrdCAs$ is transitive. Finally, let us
    assume that there exist morphisms
    $\varphi:\CAs{\gamma'}\to\CAs{\gamma}$ and
    $\psi:\CAs{\gamma}\to\CAs{\gamma'}$. In particular,
    $\psi \circ \varphi$ and $\varphi \circ \psi$ are endomorphisms of
    $\CAs{\gamma'}$ and $\CAs{\gamma}$, respectively. From
    Lemma~\ref{lem:surjective_morphisms_CAs}, these two morphisms are
    identity morphisms, so that $\varphi$ and $\psi$ are injective. From
    Lemma~\ref{lem:injective_morphisms_CAs}, $\gamma$ and $\gamma'$ are
    equal, which proves that $\OrdCAs$ is anti-symmetric. Hence,
    $\OrdCAs$ is a partial order.
\end{proof}
\medbreak

@@

Recall that $\LRank(\Tfr)$ denotes the left rank of a binary tree
$\Tfr$, defined in Section~\ref{sec:operad_Mag}. Besides, to simplify
the notation, we shall write $\bar{a}$ instead of $a - 1$ for any
integer~$a$.
\medbreak

\begin{Lemma}\label{lem:left_rank_and_CongrCAs}
    Let $\gamma$ be a positive integer and let $\Tfr$ and
    $\Tfr'$ be two binary trees. If $\Tfr \CongrCAs{\gamma} \Tfr'$, then
    \begin{equation}
        \LRank(\Tfr) \pmod{\bar{\gamma}}
        \enspace = \enspace
        \LRank\left(\Tfr'\right) \pmod{\bar{\gamma}}.
    \end{equation}
\end{Lemma}
\begin{proof}
  We consider the rewrite rule $\LComb{\gamma} \Rew \RComb{\gamma}$
    on $\Mag$. First, we show that
    \begin{equation} \label{eq:rewriting_invariant}
        \Tfr \RewContext \Tfr'
        \mbox{ implies }
        \LRank\left(\Tfr'\right) \pmod{\bar{\gamma}} = \LRank\left(\Tfr\right)
        \, \pmod{\bar{\gamma}}.
    \end{equation}
    Indeed, for all $\Tfr \in \Mag$, let
    $\Tfr_1, \dots, \Tfr_{\LRank\left(\Tfr\right)}$ be the binary trees
    such that
    \begin{math}
      \Tfr =
       \LComb{\LRank\left(\Tfr\right)} \circ
      \left[\, \Leaf \, \Tfr_1, \dots,
      \Tfr_{\LRank\left(\Tfr\right)} \right].
    \end{math}
    Thus, if $\Tfr \RewContext \Tfr'$, then one of the following two
    cases occurs:
    \begin{enumerate}[label={(\it\roman*)}]
    \item the rewrite step is applied on one of the trees $\Tfr_i$,
      that is there exists $\Tfr_i'$ such that
            \begin{math}
              \Tfr' = \LComb{\LRank\left(\Tfr\right)} \circ
              \left[\, \Leaf ,\Tfr_1, \dots, \Tfr_i',\dots,
                    \Tfr_{\LRank\left(\Tfr\right)}
              \right] \mbox{, so that }
              \LRank\left(\Tfr'\right) = \LRank\left(\Tfr\right);
            \end{math}
          \item the rewrite step is applied in the left branch beginning
            at the root, that is there exists $i$ such that
            \begin{math}
              \Tfr' = \LComb{\LRank\left(\Tfr\right)-\bar{\gamma}} \circ
              \left[ \Leaf ,\Tfr_1, \dots,
                 \RComb{\bar{\gamma}} \circ
                 [\Tfr_{i},\dots, \Tfr_{i+\bar{\gamma}}]
                 ,\dots, \Tfr_{\LRank\left(\Tfr\right)}
              \right]
              \mbox{, so that }\\
              \LRank\left(\Tfr'\right) =
              \LRank\left(\Tfr\right) - \bar{\gamma} =
              \LRank\left(\Tfr\right) \pmod{\bar{\gamma}}
            \end{math}.
    \end{enumerate}
    This concludes the proof of~\eqref{eq:rewriting_invariant}.
    \smallbreak

    We deduce from~\eqref{eq:rewriting_invariant} that the equivalence
    relation $\simeq$ defined by
    \begin{equation}
        \Tfr \simeq \Tfr' \mbox{ if and only if }
        \LRank\left(\Tfr'\right) \pmod{\bar{\gamma}} =
        \LRank\left(\Tfr\right) \, \pmod{\bar{\gamma}}
    \end{equation}
    contains the relation $\RewContext$, so that it contains its
    transitive symmetric closure. The latter is $\CongrCAs{\gamma}$,
    which concludes the proof of Lemma~\ref{lem:left_rank_and_CongrCAs}.
\end{proof}
\medbreak

@@

\Todo{Dessins}

@@

$\left(z_k \circ_{k+1} z_{n-1-k}\right)_{1 \leq k \leq n-1}$,
pictured as follows

@@

   z_k \circ_{k+1} z_{n-1-k} = dessin.

>>>>

\begin{Lemma} \label{lem:surjective_morphisms_CAs}
    Let $\gamma$ and $\gamma'$ be two positive integers. If there exists
    an operad morphism $\varphi:\CAs{\gamma'} \to \CAs{\gamma}$, then it
    is surjective and satisfies
    \begin{math}
        \varphi\left(\left[\Tfr\right]_{\CongrCAs{\gamma'}}\right)
        = \left[\Tfr\right]_{\CongrCAs{\gamma}}
    \end{math}
    for any binary tree~$\Tfr$.
\end{Lemma}
\begin{proof}
    The operad $\CAs{\gamma'}$ is generated by one binary generator
    $[\Product]_{\CongrCAs{\gamma'}}$, which is the image of the binary
    generator $\Product$ of $\Mag$ in $\CAs{\gamma'}$. Hence, $\varphi$
    is entirely determined by the image
    $\varphi\left([\Product]_{\CongrCAs{\gamma'}}\right)$. Moreover,
    $\varphi([\Product]_{\CongrCAs{\gamma'}})$ has to be of arity $2$ in
    $\CAs{\gamma}$, so that we necessarily have
    \begin{math}
        \varphi\left([\Product]_{\CongrCAs{\gamma'}}\right)
        =
        [\Product]_{\CongrCAs{\gamma}}.
    \end{math}
    Hence, if $\varphi$ exists, it is the unique operad morphism from
    $\CAs{\gamma'}$ to $\CAs{\gamma}$ determined by the image of
    $[\Product]_{\CongrCAs{\gamma'}}$. In this case,
    $[\Product]_{\CongrCAs{\gamma}}$ being in the image of $\varphi$,
    the latter is surjective. Finally, it follows that $\varphi$ sends
    $\left[\Tfr\right]_{\CongrCAs{\gamma'}}$ to
    $\left[\Tfr\right]_{\CongrCAs{\gamma}}$ by induction on the degree
    of the binary tree~$\Tfr$.
\end{proof}
\medbreak

@@

\begin{Lemma} \label{lem:morphism_CAs}
    Let $\gamma$ and $\gamma'$ be two positive integers. There exists an
    operad morphism $\varphi:\CAs{\gamma'} \to \CAs{\gamma}$ if and only
    if
    \begin{math}
      \LComb{\gamma'} \CongrCAs{\gamma} \RComb{\gamma'}.
    \end{math}
\end{Lemma}
\begin{proof}
    Assume that $\varphi:\CAs{\gamma'} \to \CAs{\gamma}$ is an operad
    morphism. Since
    \begin{math}
      \LComb{\gamma'} \CongrCAs{\gamma'} \RComb{\gamma'},
    \end{math}
    we have
    \begin{equation} \label{equ:morphism_CAs}
        \varphi\left(
        \left[\LComb{\gamma'}\right]_{\CongrCAs{\gamma'}}
        \right)
        =
        \varphi\left(
        \left[\RComb{\gamma'}\right]_{\CongrCAs{\gamma'}}
        \right).
    \end{equation}
    Now, by using Lemma~\ref{lem:surjective_morphisms_CAs}, we obtain
    from~\eqref{equ:morphism_CAs} the relation
    \begin{equation}
        \left[\LComb{\gamma'}\right]_{\CongrCAs{\gamma}}
        =
        \left[\RComb{\gamma'}\right]_{\CongrCAs{\gamma}},
    \end{equation}
    saying that
    \begin{math}
        \LComb{\gamma'} \CongrCAs{\gamma} \RComb{\gamma'}
    \end{math}
    as excepted.
    \smallbreak

@@

\begin{Proposition}\label{prop:poset_CAs}
    The binary relation $\OrdCAs$ is a partial order relation on
    $\CAsAll$.
\end{Proposition}
\begin{proof}
    The binary relation $\OrdCAs$ is reflexive since there exists the
    identity morphism on $\CAs{\gamma}$ for every positive integer
    $\gamma$. Let us now assume that
    $\varphi:\CAs{\gamma''} \to \CAs{\gamma'}$ and
    $\psi:\CAs{\gamma'} \to \CAs{\gamma}$ are two operad morphisms. By
    Lemma~\ref{lem:morphism_CAs}, one has
    $\LComb{\gamma''} \CongrCAs{\gamma'} \RComb{\gamma''}$ and
    $\LComb{\gamma'} \CongrCAs{\gamma} \RComb{\gamma'}$. This implies
    that $\CongrCAs{\gamma}$ is coarser than $\CongrCAs{\gamma'}$ and
    that $\CongrCAs{\gamma'}$ is coarser that $\CongrCAs{\gamma''}$. Now,
    again by Lemma~\ref{lem:morphism_CAs}, $\psi \circ \varphi$ is an
    operad morphism. Hence, $\OrdCAs$ is transitive. Finally, let us
    assume that there exist morphisms
    $\varphi:\CAs{\gamma'}\to\CAs{\gamma}$ and
    $\psi:\CAs{\gamma}\to\CAs{\gamma'}$. In particular,
    $\psi \circ \varphi$ and $\varphi \circ \psi$ are endomorphisms of
    $\CAs{\gamma'}$ and $\CAs{\gamma}$, respectively. From
    Lemma~\ref{lem:surjective_morphisms_CAs}, these two morphisms are
    identity morphisms, so that $\varphi$ and $\psi$ are injective. From
    Lemma~\ref{lem:injective_morphisms_CAs}, $\gamma$ and $\gamma'$ are
    equal, which proves that $\OrdCAs$ is anti-symmetric. Hence,
    $\OrdCAs$ is a partial order.
\end{proof}
\medbreak

@@

Recall that $\LRank(\Tfr)$ denotes the left rank of a binary tree
$\Tfr$, as defined in Section~\ref{sec:operad_Mag}. Besides, to simplify
the notation, we shall write $\bar{a}$ instead of $a - 1$ for any
integer~$a$.
\medbreak

\begin{Lemma}\label{lem:left_rank_and_CongrCAs}
    Let $\gamma$ be a positive integer and let $\Tfr$ and $\Tfr'$ be two
    binary trees. If $\Tfr \CongrCAs{\gamma} \Tfr'$, then
    \begin{equation} \label{equ:left_rank_and_CongrCAs}
        \LRank(\Tfr) \pmod{\bar{\gamma}}
        \enspace = \enspace
        \LRank\left(\Tfr'\right) \pmod{\bar{\gamma}}.
    \end{equation}
\end{Lemma}
\begin{proof}
    Consider here the rewrite rule $\Rew$ on $\Mag$ satisfying
    $\LComb{\gamma} \Rew \RComb{\gamma}$. Let us show that the rewrite
    relation $\RewContext$ induced by $\Rew$ is such that
    $\Tfr \RewContext \Tfr'$ implies~\eqref{equ:left_rank_and_CongrCAs}.
    Any binary tree $\Tfr$ decomposes as
    \begin{math}
        \Tfr =
        \LComb{\LRank\left(\Tfr\right)} \circ
        \left[\Leaf, \Tfr_1, \dots,
        \Tfr_{\LRank\left(\Tfr\right)} \right]
    \end{math}
    where the $\Tfr_i$ are binary trees. Now, if
    $\Tfr \RewContext \Tfr'$, then one among the following two cases
    occurs.
    \begin{enumerate}[label={(\it\roman*)}]
        \item The rewrite step is applied into one of the trees
        $\Tfr_i$, that is there exists $\Tfr_i'$ such that
        \begin{math}
            \Tfr' = \LComb{\LRank\left(\Tfr\right)}
            \circ \left[\Leaf, \Tfr_1, \dots, \Tfr_i', \dots,
                \Tfr_{\LRank\left(\Tfr\right)}
            \right]
        \end{math}
        so that
        \begin{math}
            \LRank\left(\Tfr'\right) = \LRank\left(\Tfr\right).
        \end{math}
        \item The rewrite step is applied into the left branch
          beginning
            at the root of $\Tfr$, that is there exists $i$ such that
            \begin{math}
                \Tfr' = \LComb{\LRank\left(\Tfr\right)-\bar{\gamma}}
                \circ \left[\Leaf, \Tfr_1, \dots, \RComb{\bar{\gamma}}
                \circ
                \left[\Tfr_{i},\dots, \Tfr_{i+\bar{\gamma}}\right],
                \dots, \Tfr_{\LRank\left(\Tfr\right)} \right]
            \end{math}
            so that
            \begin{math}
                \LRank\left(\Tfr'\right) =
                \LRank\left(\Tfr\right) - \bar{\gamma} =
                \LRank\left(\Tfr\right) \pmod{\bar{\gamma}}
            \end{math}.
    \end{enumerate}
    This implies~\eqref{equ:left_rank_and_CongrCAs}. Finally, since
    $\CongrCAs{\gamma}$ is the reflexive, symmetric, and transitive
    closure of~$\RewContext$, the statement of the lemma follows.
\end{proof}
\medbreak

@@

@@

$\left(z_k \circ_{k+1} z_{n-1-k}\right)_{1 \leq k \leq n-1}$.
For example,

@@

   z_8 \circ_{9} z_{3} =
   \begin{tikzpicture}[xscale=.18,yscale=.16,Centering]
      \node(0)at(0.00,-5.56){};
      \node(10)at(10.00,-19.44){};
      \node(12)at(12.00,-13.89){};
      \node(14)at(14.00,-8.33){};
      \node(16)at(16.00,-8.33){};
      \node(18)at(18.00,-13.89){};
      \node(2)at(2.00,-11.11){};
      \node(20)at(20.00,-13.89){};
      \node(22)at(22.00,-11.11){};
      \node(24)at(24.00,-5.56){};
      \node(4)at(4.00,-16.67){};
      \node(6)at(6.00,-22.22){};
      \node(8)at(8.00,-22.22){};
      \node(19)at(19.00,-11){};
      \node[NodeST](1)at(1.00,-2.78){\begin{math}\Product\end{math}};
      \node[NodeST](11)at(11.00,-11.11){\begin{math}\Product\end{math}};
      \node[NodeST](13)at(13.00,-5.56){\begin{math}\Product\end{math}};
      \node[NodeST](15)at(15.00,0.00){\begin{math}\Product\end{math}};
      \node[NodeST](17)at(17.00,-5.56){\begin{math}\Product\end{math}};
      \node[NodeST](21)at(21.00,-8.33){\begin{math}\Product\end{math}};
      \node[NodeST](23)at(23.00,-2.78){\begin{math}\Product\end{math}};
      \node[NodeST](3)at(3.00,-8.33){\begin{math}\Product\end{math}};
      \node[NodeST](5)at(5.00,-13.89){\begin{math}\Product\end{math}};
      \node[NodeST](7)at(7.00,-19.44){\begin{math}\Product\end{math}};
      \node[NodeST](9)at(9.00,-16.67){\begin{math}\Product\end{math}};
      \draw[Edge](1)--(15);
      \draw[Edge](11)--(3);
      \draw[Edge](13)--(1);
      \draw[Edge](16)--(17);
      \draw[Edge](17)--(23);
      \draw[Edge](19)--(21);
      \draw[Edge](21)--(17);
      \draw[Edge](22)--(21);
      \draw[Edge](23)--(15);
      \draw[Edge](24)--(23);
      \draw[Edge](3)--(13);
      \draw[Edge](5)--(11);
      \draw[Edge](9)--(5);
      \node(r)at(15.00,2.08){};
      \draw[Edge](r)--(15);
